Crowdsourcing and phylogenetic modelling reveal parrot tool use is not rare

### 2023-11-20 manuscript.Rmd
prepare for submission

abstract: |
 Studying the prevalence of putatively rare behaviours, such as tool use, is challenging because absence of evidence can arise either from a species' inability to produce the behaviour or from insufficient research effort. Here, we tackle this challenge by combining crowdsourcing and phylogenetic modelling to approximate actual rates of a rarely observed behaviour based on limited data, targeting tool use in parrots. Crowdsourcing on a social media platform revealed novel instances of tool use in 17 parrot species, more than doubling the confirmed number of tool-using parrot species from 11 (3%) to 28 (7%). Phylogenetic modelling ranked additional species that are most likely to be unobserved tool users, suggesting that between 11% and 17% of extant parrot species may be tool users. These discoveries have implications for inferences about the evolutionary drivers and origins of tool use in parrots, revealing associations with relative brain size and feeding generalism and indicating several genera where tool use was likely an ancestral trait. Overall, our findings challenge the assumption that current sampling effort captures the full distribution of putatively rare animal behaviours. Combining our sampling and analysis methods offers a fruitful approach for investigating the distribution, drivers, and origins of other rare behaviours.

keywords          : "tool use; technical intelligence; parrots; crowdsourcing; phylogenetic modelling"
wordcount         : "4289 words excluding Methods (Methods: 1821 words)"
bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
header-includes: 
  - \usepackage{array}
  - \usepackage{caption}
  - \usepackage{graphicx}
  - \usepackage{siunitx}
  - \usepackage[normalem]{ulem}
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{hhline}
  - \usepackage{calc}
  - \usepackage{tabularx}
  - \usepackage{threeparttable}
  - \usepackage{wrapfig}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \raggedbottom
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1,small}}
  - \captionsetup[figure]{font={stretch=1,small}}
  - \nolinenumbers
  - \note{\raggedright † indicates shared first authorship, ‡ indicates shared senior authorship \par * Correspondence concerning this article should be addressed to Scott Claessens, Level 2, Building 302, 23 Symonds Street, Auckland, New Zealand. \text{E-mail:} \href{scott.claessens@gmail.com}{\nolinkurl{scott.claessens@gmail.com}} \par This working paper has not yet been peer-reviewed.}
appendix          : "appendix.Rmd"
csl               : nature.csl
---

\linenumbers

Our understanding of the evolution of animal behaviour is built on the assumption that we have access to sufficient data[@MacLean2012; @Lefebvre2011; @Logan2018]. However, this is not always the case. Data on behaviours that are rare, fleeting, or otherwise difficult to observe are likely to be patchy and incomplete[@Healy2007; @Powell2017]. Among species for which such behaviours have not been observed, it can be difficult to differentiate between cases in which the species is truly incapable of producing the behaviour and cases in which the species is capable of producing the behaviour but the behaviour has not yet been observed. Such a distinction can be critical for drawing conclusions about the rarity and evolution of the behaviour in question.

Comparative work on the evolution of tool use is a paradigmatic example of this issue. The initial discoveries of tool use in chimpanzees[@Goodall1964], birds[@Hunt1996], dolphins[@Smolker1997], and octopuses[@Finn2009] occurred decades after significant advances on other more easily measurable aspects of their biology. Since these initial discoveries, scholars have used the distribution of reported tool use across species to make various claims about the evolutionary drivers of tool use behaviours. For example, based on the observation that bird species with reported tool use tended to have larger brains, researchers have identified higher relative brain size as a likely precondition for tool using capabilities[@Lefebvre2002;@Lefebvre1997;@Lefebvre2004]. These researchers argue that larger brains are better able to integrate visual and somatosensory information when innovating novel behaviours, such as tool use, in changing environments[@Sol2005;@Sol2009]. Similarly, researchers have used existing reports of tool use in birds to debate the roles of generalist versus specialist feeding strategies in driving the evolution of tool use, with some arguing that feeding generalists require technical innovations to expand their dietary niche[@Ducatez2015;@Overington2011;@Sol2005] and others arguing that feeding specialists require technical innovations for extractive foraging of specific foods[@HenkeVonDerMalsburg2020;@MettkeHofmann2014].

However, before we can make claims of this kind, we need to know whether current research effort in the literature is sufficient for robust conclusions to be drawn about the evolution of tool use. In fact, evidence suggests that research effort is often systematically biased towards particular taxonomic groups, parts of the world that are easy to access, and species with life history traits that make them easier to study, such as larger distribution ranges and population sizes[@Ducatez2014]. This is a crucial limitation because insufficient observation may lead researchers to miss true instances of tool behaviours and thus draw premature conclusions about the evolutionary drivers and origins of tool use. While researchers have attempted to deal with this problem by controlling for the number of scientific papers published on different species, previous work has not yet attempted to quantify and explicitly model the relationship between actual tool-using behaviour and what is reported in the scientific literature. If more tool-using species exist than previously thought, this could have important implications for theories of the evolutionary drivers and origins of tool use and, more generally, for our understanding of how rare this behaviour actually is.

One potentially powerful method for quantifying actual rates of rare animal behaviours is crowdsourcing[@Nelson2013]. In a crowdsourcing study, researchers collect reports from the general public and/or collate and analyse videos posted on social media platforms. This citizen science approach has been widely used in ecology to monitor the distributional patterns of species[@Dickinson2010], but has also recently been used to uncover a variety of rare animal behaviours, including interspecies play in dogs[@Nelson2013], novel problem-solving behaviours in horses[@Krueger2019], and socially-learned foraging innovations in cockatoos[@Klump2021;@Klump2022]. By casting the net wider than the scientific literature, the crowdsourcing method can provide an indication of the tool-using species that the literature might be missing.

Even after using this crowdsourcing approach, some tool-users could _still_ remain unobserved. An additional approach to identify these unobserved species is to specify a causal model of the process that generates the observed data. We propose one such causal model in Figure \@ref(fig:plotDAG). In this model, we assume that the presence of tool use in the scientific literature (or in crowdsourced reports) is caused by both unobserved tool use capabilities and the number of published studies (or the number of crowdsourced reports) for any given species. Tool users are more likely to be observed if they are well studied, but understudied tool users may go undetected. Furthermore, based on existing theories of the evolution of tool use[@Ducatez2015;@HenkeVonDerMalsburg2020;@Lefebvre1997;@Lefebvre2002;@Lefebvre2004;@MettkeHofmann2014;@Overington2011;@Sol2005;@Sol2009], we propose that the unobserved presence or absence of tool use is additionally caused by relative brain size, feeding strategy, and shared phylogenetic ancestry. Expressing this causal model as a statistical model can suggest further species which are likely to be unobserved tool-users and, simultaneously, test existing theories of the evolutionary drivers of tool use without incorrectly assuming that absence of evidence is evidence of absence.

@@

Here, we apply these crowdsourcing and phylogenetic modelling approaches to tool use in the parrot order. We focus on tool use in parrots for a number of reasons. First, the scientific literature suggests that only a small proportion of extant parrot species (11 out of 398[@Burgio2019]; 3%) use tools[@Auersperg2009;@Auersperg2010;@Auersperg2011a;@Auersperg2011b;@Auersperg2012;@Bastos2021;@BentleyCondit2010;@Borsari2005;@Goodman2018;@Heinsohn2017;@Janzen1976;@Lambert2015;@Lefebvre2002;@OHara2021] (note that we do not include in this count species that have been previously shown to use touchscreen devices as these behaviours were explicitly trained[@Kleinberger2023]). Parrot tool use thus provides an ideal test case for examining how robust sampling is in the scientific literature. Second, parrots are highly popular as pets. Over 70% of all extant parrot species are bred in the aviculture industry and kept as pets worldwide[@Anderson2003;@Carrete2008;@Drews2001;@Kelly2014;@Li2014;@Su2015;@Young2012], enabling us to leverage the power of crowdsourcing on a social media platform to search for evidence of tool use[@Nelson2013]. Third, detailed data on relative brain sizes[@Flammer2001;@Iwaniuk2005;@Mazengenya2018;@Olkowicz2016;@Schuck2008;@Silva2017], feeding strategies[@Wilman2016], and shared ancestry[@Jetz2012] exist for parrots, allowing us to fit the statistical model implied by Figure \@ref(fig:plotDAG) to the entire parrot order.

@@

## Crowdsourcing video survey

We surveyed the social media platform YouTube for video evidence of tool use in parrots (see Methods for detailed search criteria). In our search, we defined "true" tool use behaviour as the manipulation of an unattached object to achieve a goal[@Shumaker2011], while "borderline" tool use involved the use of an object that was still attached to a substrate[@Seed2010].

@@

In total, we found `r nrow(crowdData)` videos of `r crowdData %>% pull(IndividualID) %>% unique() %>% length()` individuals from `r crowdData %>% filter(Species != "Unknown") %>% pull(Species) %>% unique() %>% length()` parrot species performing either true tool use (`r nrow(trueCrowdData)` videos of `r trueCrowdData %>% pull(IndividualID) %>% unique() %>% length()` individuals from `r trueCrowdData %>% filter(Species != "Unknown") %>% pull(Species) %>% unique() %>% length()` species) or borderline tool use (`r nrow(bordCrowdData)` videos of `r bordCrowdData %>% pull(IndividualID) %>% unique() %>% length()` individuals from `r bordCrowdData %>% filter(Species != "Unknown") %>% pull(Species) %>% unique() %>% length()` species). All videos featured pet parrots in captive settings. In `r crowdData %>% pull(HumanInfluence) %>% str_starts("No") %>% sum()` of these videos, owners did not appear to interfere with the subjects' actions. In `r crowdData %>% pull(HumanInfluence) %>% str_starts("Yes") %>% sum()` videos, there was potential interference, either from the owners being in close physical contact with the bird (e.g., bird perching on hand), talking to the bird, or handing it the tool (which occurred in only two videos). We could not establish the degree of interference in the remaining `r crowdData %>% pull(HumanInfluence) %>% str_starts("Unknown") %>% sum()` videos, as sound had been removed or was substituted by music. None of the videos featured owners directly rewarding tool use behaviours with food. All borderline tool use cases were excluded from further analyses.

@@

According to YouTube video descriptions and owner comments, `r trueCrowdData %>% group_by(IndividualID, Sex) %>% summarise(.groups = "drop") %>% pull(Sex) %>% (function(x) sum(x == "Male"))` of the individuals performing true tool use were males and `r trueCrowdData %>% group_by(IndividualID, Sex) %>% summarise(.groups = "drop") %>% pull(Sex) %>% (function(x) sum(x == "Female"))` were females. No sex information was provided for the remaining `r trueCrowdData %>% group_by(IndividualID, Sex) %>% summarise(.groups = "drop") %>% pull(Sex) %>% (function(x) sum(x == "Unknown"))` individuals. As owners provided no information on whether sex had been established through genetic testing, and sexual dimorphism in parrots is rare[@Bercovitz1987;@Hoyo2011], we could not typically ascertain if descriptions were accurate. It is unclear if the disproportionately large number of males in the sample is a consequence of owners more readily assuming their parrots are male when they have not been genetically tested, owners being more likely to own or film male parrots, or male parrots exhibiting more true tool use behaviours than female parrots.

Figure \@ref(fig:plotPhylo1) maps the findings from the video survey onto a maximum clade credibility phylogeny for the parrot order, plotted alongside species previously identified in the scientific literature. Before the video survey, 11 parrot species (3%) had been identified as tool users in the scientific literature. Across our video survey, we observed true tool use in `r trueCrowdData %>% filter(Species != "Unknown") %>% pull(Species) %>% unique() %>% length()` species, 5 of which overlapped with the scientific literature and `r (trueCrowdData %>% filter(Species != "Unknown") %>% pull(Species) %>% unique() %>% length()) - 5` of which were novel species. All of the species identified in the video survey were cockatoos (Cacatuidae), Old World parrots (Psittacinae), or neotropical parrots (Arinae). The most common species in our survey, accounting for `r crowdData %>% filter(Species == "Pyrrhura molinae") %>% nrow()` videos from `r crowdData %>% filter(Species == "Pyrrhura molinae") %>% pull(IndividualID) %>% unique() %>% length()` individuals, was the green-cheeked conure (_Pyrrhura molinae_). In accordance with the scientific literature, the video survey found no evidence of tool use in any species of Psittaculidae, despite this family containing some of the most commonly kept pet species, including lovebirds, lorikeets, and Asian parakeet species. Combining both the video survey and the scientific literature, we can thus identify 28 tool-using parrot species overall (7%), compared to the 11 previously reported.

(ref:plotPhylo1Caption) _Results of crowdsourcing video survey and phylogenetic survival cure modelling mapped onto a maximum clade credibility phylogeny of the parrot order._ Purple tips indicate species observed in the scientific literature and in the video survey. Blue tips indicate species observed only in the scientific literature. Orange tips indicate species observed only in the video survey (note that three species observed only in the video survey are not present in the phylogeny due to a lack of genomic data: _Psittacara erythrogenys_, _Psittacus timneh_, and _Aratinga nenday_). Yellow tips indicate the top ten most likely tool-using species from our phylogenetic survival cure model which were not observed in the scientific literature or the video survey. All other tips are labelled as absent (grey). Total _n_ = `r nrow(tar_read(d))` species.

@@

## Phylogenetic survival cure model

@@

Survival cure models are often used in the medical sciences to analyse the time to some event, such as cancer relapse, with censored patient data[@Amico2018]. The data are right-censored because some patients have relapsed when they are measured, and some have not yet relapsed when they are measured (i.e., they are censored). The cure aspect of these models comes from an additional assumption that a certain proportion of the population is "cured" and can never relapse, no matter how long we measure them for.

Our tool use problem has the same features. We are modelling a time-to-event; specifically, the amount of "time" (i.e., observation opportunities measured as the number of published papers or crowdsourced videos) until tool use is identified. This is right-censored data, because many species will not have had tool use identified when we measure them. Moreover, we can assume that a certain proportion of the population is "cured" -- that is, they are not tool users, and so we will never identify tool use for them, no matter how long we measure them for.

In our model, we infer the tool-using status of each species by allowing each species to have their own probability of being "cured" (i.e., not a tool user). Following our causal model (Figure \@ref(fig:plotDAG)), we predict these probabilities based on feeding strategy, encephalisation quotient, and phylogenetic history (see Methods for full model). The model additionally takes research effort into account by allowing that, among species for which tool use is unobserved, all else being equal those with fewer published papers and fewer video search hits have a higher probability of being undetected tool users (Supplementary Figure \@ref(fig:plotSurvCure3)).

@@

We found that this phylogenetic survival cure model was able to adequately distinguish between species with and without evidence for tool use, with an area-under-the-curve classification statistic of `r printnum(tar_read(auc), digits = 2)` (Supplementary Figure \@ref(fig:plotSurvCure9)). To further estimate the accuracy of the model's predictions, we also used a leave-one-species-out approach with known tool users. For each of the 25 tool-using species that were represented on the phylogeny and for which we had brain size and genomic data (we lacked data for three tool-using species), we fitted the model to a modified dataset which set tool use to be absent for the target species in both the scientific literature and the video survey. Across `r length(tar_read(cv))` cross-validation models, `r sum(tar_read(cv))` models (`r printnum(mean(tar_read(cv))*100, digits = 0)`%) continued to predict the target species as having a median posterior probability of tool use that was within the range of all other tool users. This classification rate was greater than the baseline classification rate of `r printnum(mean(plotSurvCure1$data %>% filter(ToolUse == "Absent") %>% pull(med) > 0.16478274) * 100, digits = 0)`% for species without evidence of tool use in the full model (`r sum(plotSurvCure1$data %>% filter(ToolUse == "Absent") %>% pull(med) > 0.16478274)` of `r nrow(plotSurvCure1$data %>% filter(ToolUse == "Absent"))` species without evidence of tool use had a median posterior probability of tool use that was within the range of the tool-using species). Together, the area-under-the-curve statistic and the leave-one-species-out approach suggest that the model is able to adequately classify known tool users, with some error.

>>>>

abstract: |
 Putatively rare behaviours, such as tool use, are challenging to study because absence of evidence can arise either from a species’ inability to produce the behaviour or from insufficient research effort. Here, we tackle this challenge by combining crowdsourcing and phylogenetic modelling to estimate actual rates of tool use in parrots. Crowdsourcing revealed novel instances of tool use in 17 parrot species, more than doubling the confirmed number of tool-using parrot species from 11 (3%) to 28 (7%). Phylogenetic models incorporating these data together with measures of research effort and hypothesised predictors of tool use suggest that between 11% and 17% of parrot species may be tool users. These findings challenge the assumption that current research effort captures the distribution of putatively rare animal behaviours.

keywords          : "tool use; technical intelligence; parrots; crowdsourcing; phylogenetic modelling"
wordcount         : "6548 words"
bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
header-includes: 
  - \usepackage{array}
  - \usepackage{caption}
  - \usepackage{graphicx}
  - \usepackage{siunitx}
  - \usepackage[normalem]{ulem}
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{hhline}
  - \usepackage{calc}
  - \usepackage{tabularx}
  - \usepackage{threeparttable}
  - \usepackage{wrapfig}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \raggedbottom
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1,small}}
  - \captionsetup[figure]{font={stretch=1,small}}
  - \nolinenumbers
  - \note{\raggedright † These authors contributed equally to this work \par ‡ These authors contributed equally to this work \par * Correspondence concerning this article should be addressed to Scott Claessens, Level 2, Building 302, 23 Symonds Street, Auckland, New Zealand. \text{E-mail:} \href{scott.claessens@gmail.com}{\nolinkurl{scott.claessens@gmail.com}} \par This working paper has not yet been peer-reviewed.}
appendix          : "appendix.Rmd"
csl               : science.csl
---

\nolinenumbers

Our understanding of the evolution of animal behaviour is built on the 
assumption that we have access to sufficient data [@MacLean2012; @Lefebvre2011; 
@Logan2018]. However, this is not always the case. Data on behaviours that are 
rare, fleeting, or otherwise difficult to observe are likely to be patchy and 
incomplete [@Healy2007; @Powell2017]. Among species for which such behaviours 
have not been observed, it can be difficult to differentiate between cases in 
which the species is truly incapable of producing the behaviour and cases in 
which the species is capable of producing the behaviour but the behaviour has 
not yet been observed. Such a distinction can be critical for drawing 
conclusions about the rarity and evolution of the behaviour in question.

Comparative work on the evolution of tool use is a paradigmatic example of this 
issue. The initial discoveries of tool use in chimpanzees [@Goodall1964], 
birds [@Hunt1996], dolphins [@Smolker1997], and octopuses [@Finn2009] occurred 
decades after significant advances on other more easily measurable aspects of 
their biology. Since then, scholars have proposed a clear operational 
definition of tool use applicable to all species --- the manipulation of an 
unattached object as an extension of the animal's body to achieve a 
goal [@Shumaker2011] --- and have used the distribution of species meeting this 
definition to make various claims about the evolutionary drivers of tool use 
behaviours. For example, based on the observation that bird species with 
reported tool use tended to have larger brains, researchers have identified 
higher relative brain size as a likely precondition for tool using 
capabilities [@Lefebvre2002;@Lefebvre1997;@Lefebvre2004]. These researchers 
argue that larger brains are better able to integrate visual and somatosensory 
information when innovating novel behaviours, such as tool use, in changing 
environments [@Sol2005;@Sol2009]. Similarly, researchers have used existing 
reports of tool use in birds to debate the roles of generalist versus 
specialist feeding strategies in driving the evolution of tool use, with some 
arguing that feeding generalists require technical innovations to expand their 
dietary niche [@Ducatez2015;@Overington2011;@Sol2005] and others arguing that 
feeding specialists require technical innovations for extractive foraging of 
specific foods [@HenkeVonDerMalsburg2020;@MettkeHofmann2014].

However, before we can make claims of this kind, we need to know whether current 
research effort in the literature is sufficient for robust conclusions to be 
drawn about the evolution of tool use. In fact, evidence suggests that research 
effort is often systematically biased towards particular taxonomic groups, parts 
of the world that are easy to access, and species with life history traits that 
make them easier to study, such as larger distribution ranges and population 
sizes [@Ducatez2014]. This is a crucial limitation because insufficient 
observation may lead researchers to miss true instances of tool behaviours and 
thus draw premature conclusions about the evolutionary drivers and origins of 
tool use. While researchers have attempted to deal with this problem by 
controlling for the number of scientific papers published on different species, 
previous work has not yet attempted to quantify and explicitly model the 
relationship between actual tool-using behaviour and what is reported in the 
scientific literature. If more tool-using species exist than previously thought, 
this could have important implications for theories of the evolutionary drivers 
and origins of tool use and for our understanding of how rare this behaviour 
actually is.

One potentially powerful method for quantifying actual rates of rare animal 
behaviours is crowdsourcing [@Nelson2013]. In a crowdsourcing study, researchers 
collect reports from the general public and/or collate and analyse videos posted 
on social media platforms. This citizen science approach has been widely used in 
ecology to monitor the distributional patterns of species [@Dickinson2010], but 
has also recently been used to uncover a variety of rare animal behaviours, 
including interspecies play in dogs [@Nelson2013], novel problem-solving 
behaviours in horses [@Krueger2019], and socially-learned foraging innovations 
in cockatoos [@Klump2021;@Klump2022]. By casting the net wider than the 
scientific literature, the crowdsourcing method can provide an indication of the 
tool-using species that the literature might be missing.

Even after using this crowdsourcing approach, some tool-users could _still_ 
remain unobserved. One principled framework for identifying these unobserved species 
is to specify a causal model of the process that generates the observed data. We 
propose one such causal model in Figure \@ref(fig:plotDAG). In this model, we 
assume that the presence of tool use in the scientific literature (or in 
crowdsourced reports) is caused by both unobserved tool use capabilities and the 
number of published studies (or the number of crowdsourced reports) for any 
given species. Tool users are more likely to be observed if they are well 
studied, but understudied tool users may go undetected. Furthermore, based on 
existing theories of the evolution of tool use [@Ducatez2015;
@HenkeVonDerMalsburg2020;@Lefebvre1997;@Lefebvre2002;@Lefebvre2004;
@MettkeHofmann2014;@Overington2011;@Sol2005;@Sol2009], we propose that the 
unobserved tool use capabilities are additionally caused by relative brain size, 
feeding strategy, and shared phylogenetic ancestry. Expressing this causal model 
as a statistical model can suggest further species which are likely to be 
unobserved tool-users and, simultaneously, test existing theories of the 
evolutionary drivers of tool use without incorrectly assuming that absence of 
evidence is evidence of absence.

@@

Here, we apply these crowdsourcing and phylogenetic modelling approaches to tool 
use in the parrot order. We focus on tool use in parrots for a number of 
reasons. First, the scientific literature suggests that only a small proportion 
of extant parrot species (11 out of 398; 3%) use tools 
[@Auersperg2009;@Auersperg2010;@Auersperg2011a;@Auersperg2011b;@Auersperg2012;
@Bastos2021;@BentleyCondit2010;@Borsari2005;@Goodman2018;@Heinsohn2017;
@Janzen1976;@Lambert2015;@Lefebvre2002;@OHara2021]. Parrot tool use 
thus provides an ideal test case for examining how robust sampling is in the 
scientific literature. Second, parrots are highly popular as pets. Over 70% of 
all extant parrot species are bred in the aviculture industry and kept as pets 
worldwide [@Anderson2003;@Carrete2008;@Drews2001;@Kelly2014;@Li2014;@Su2015;
@Young2012], enabling us to leverage the power of crowdsourcing on a social 
media platform to search for evidence of tool use [@Nelson2013]. Third, detailed 
data on relative brain sizes [@Flammer2001;@Iwaniuk2005;@Mazengenya2018;
@Olkowicz2016;@Schuck2008;@Silva2017], feeding strategies [@Wilman2016], and 
shared ancestry [@Jetz2012] exist for parrots, allowing us to fit the 
statistical model implied by Figure \@ref(fig:plotDAG) to the entire parrot 
order.

@@

## Crowdsourcing reveals tool use in additional parrot species

We surveyed the social media platform YouTube for video evidence of tool use in 
parrots (see Methods for detailed search criteria). In our search, we used the 
standard criteria for identifying tool use in the literature, defining "true" 
tool use behaviour as the manipulation of an unattached object as an extension 
of the animal's body to achieve a goal [@Shumaker2011], while "borderline" tool 
use involved the use of an object that was still attached to a substrate
[@Seed2010].

@@

In total, we found `r nrow(crowdData)` videos of 
`r crowdData %>% pull(IndividualID) %>% unique() %>% length()` individuals from 
`r crowdData %>% filter(Species != "Unknown") %>% pull(Species) %>% unique() %>% length()` 
parrot species performing behaviours that met the definition of either true tool 
use (`r nrow(trueCrowdData)` videos of 
`r trueCrowdData %>% pull(IndividualID) %>% unique() %>% length()` 
individuals from 
`r trueCrowdData %>% filter(Species != "Unknown") %>% pull(Species) %>% unique() %>% length()` 
species) or borderline tool use (`r nrow(bordCrowdData)` videos of 
`r bordCrowdData %>% pull(IndividualID) %>% unique() %>% length()` 
individuals from 
`r bordCrowdData %>% filter(Species != "Unknown") %>% pull(Species) %>% unique() %>% length()` 
species). All videos featured pet parrots in captive settings. In 
`r crowdData %>% pull(HumanInfluence) %>% str_starts("No") %>% sum()` of these 
videos, owners did not appear to interact with the subjects. In 
`r crowdData %>% pull(HumanInfluence) %>% str_starts("Yes") %>% sum()` videos, 
there was potential human interaction, either from the owners being in close physical 
contact with the bird (e.g., bird perching on hand), talking to the bird, or 
handing it the tool (which occurred in only two videos). We could not establish 
the degree of human interaction in the remaining 
`r crowdData %>% pull(HumanInfluence) %>% str_starts("Unknown") %>% sum()` 
videos, as sound had been removed or was substituted by music. None of the 
videos featured owners directly rewarding tool use behaviours with food. All 
borderline tool use cases were excluded from further analyses.

@@

According to YouTube video descriptions and owner comments, 
`r trueCrowdData %>% group_by(IndividualID, Sex) %>% summarise(.groups = "drop") %>% pull(Sex) %>% (function(x) sum(x == "Male"))` 
of the individuals performing true tool use were males and 
`r trueCrowdData %>% group_by(IndividualID, Sex) %>% summarise(.groups = "drop") %>% pull(Sex) %>% (function(x) sum(x == "Female"))` 
were females. No sex information was provided for the remaining 
`r trueCrowdData %>% group_by(IndividualID, Sex) %>% summarise(.groups = "drop") %>% pull(Sex) %>% (function(x) sum(x == "Unknown"))` 
individuals. As owners provided no information on whether sex had been 
established through genetic testing, and sexual dimorphism in parrots is 
rare [@Bercovitz1987;@Hoyo2011], we could not typically ascertain if descriptions 
were accurate. It is unclear if the disproportionately large number of males in 
the sample is a consequence of owners more readily assuming their parrots are 
male when they have not been genetically tested, owners being more likely to own 
or film male parrots, or male parrots exhibiting more true tool use behaviours 
than female parrots.

Figure \@ref(fig:plotPhylo1) maps the findings from the video survey onto a 
maximum clade credibility phylogeny for the parrot order, plotted alongside 
species previously identified in the scientific literature. Before the video 
survey, 11 parrot species (3%) had been identified as tool users in the 
scientific literature. Across our video survey, we observed true tool use in 
`r trueCrowdData %>% filter(Species != "Unknown") %>% pull(Species) %>% unique() %>% length()` 
species, 5 of which overlapped with the scientific literature and 
`r (trueCrowdData %>% filter(Species != "Unknown") %>% pull(Species) %>% unique() %>% length()) - 5` 
of which were novel species. All of the species identified in the video survey 
were cockatoos (_Cacatuidae_), Old World parrots (_Psittacinae_), or neotropical 
parrots (_Arinae_). The most common species in our survey, accounting for 
`r crowdData %>% filter(Species == "Pyrrhura molinae") %>% nrow()` videos from 
`r crowdData %>% filter(Species == "Pyrrhura molinae") %>% pull(IndividualID) %>% unique() %>% length()` 
individuals, was the green-cheeked conure (_Pyrrhura molinae_). In accordance 
with the scientific literature, the video survey found no evidence of tool use 
in any species of Psittaculidae, despite this family containing some of the most 
commonly kept pet species, including lovebirds, lorikeets, and Asian parakeet 
species. Combining both the video survey and the scientific literature, we can 
thus identify 28 tool-using parrot species overall (7%), compared to the 11 
previously reported.

(ref:plotPhylo1Caption) _Results of crowdsourcing video survey and phylogenetic 
survival cure modelling mapped onto a maximum clade credibility phylogeny of the 
parrot order._ Orange points in the inner ring indicate species observed in the 
video survey, with point size scaled by the number of videos for each species 
(note that three species observed only in the video survey are not present in 
the phylogeny due to a lack of genomic data: _Psittacara erythrogenys_, 
_Psittacus timneh_, and _Aratinga nenday_). Blue points in the outer ring 
indicate species observed in the scientific literature, with point size scaled 
by the number of papers published on each species. Yellow species names indicate 
the top ten most likely tool-using species from our phylogenetic survival cure 
model which were not observed in the scientific literature or the video survey. 
Total _n_ = `r nrow(tar_read(d))` species.

@@

## Phylogenetic survival cure modelling identifies further candidate tool users

@@

Survival cure models [@Amico2018], also known as split population 
models [@Schmidt1989], are used to analyse the time to some event of interest 
with the added assumption that a certain proportion of the population will never 
experience the event, no matter how long they are measured for. These models 
have been used to analyse a variety of right-censored outcomes, from cancer 
relapse [@Amico2018] to criminal recidivism [@Schmidt1989]. The data are 
right-censored because some individuals will have experienced the event when 
they are measured (e.g., disease onset, return to prison) while others will have 
not experienced the event. For those who have not experienced the event, this 
may be because (_i_) the event has not happened to them yet or (_ii_) the event 
will _never_ happen to them. Survival cure models treat these two processes 
separately.

Our tool use problem has the same features. We are modelling a time-to-event; 
specifically, the amount of "time" (i.e., observation opportunities measured as 
the number of published papers or crowdsourced videos) until tool use is 
identified. This is right-censored data because many species will not have had 
tool use identified when we measure them. Moreover, we can assume that a certain 
proportion of the population will never experience the event -- that is, they 
are a non-tool-user and so we will never identify tool use no matter how long we 
measure them for.

In our model, we infer the tool-using status of each species by allowing each 
species to have their own probability of being a non-tool-user. Following our 
causal model (Figure \@ref(fig:plotDAG)), we predict these probabilities based 
on feeding strategy, encephalisation quotient, and phylogenetic history (see 
Methods for full model). The model additionally takes research effort into 
account by allowing that, among species for which tool use is unobserved, all 
else being equal those with fewer published papers and fewer video search hits 
have a higher probability of being undetected tool users (fig. 
\@ref(fig:plotSurvCure3)).

@@

We found that this phylogenetic survival cure model was able to adequately 
distinguish between species with and without evidence for tool use, with an 
area-under-the-curve classification statistic of 
`r printnum(tar_read(auc), digits = 2)` (fig.  
\@ref(fig:plotSurvCure9)). To further estimate the accuracy of the model's 
predictions, we also used a leave-one-species-out approach with known tool 
users. For each of the 25 tool-using species that were represented on the 
phylogeny and for which we had brain size and genomic data (we lacked data for 
three tool-using species), we fitted the model to a modified dataset which set 
tool use to be absent for the target species in both the scientific literature 
and the video survey. Across `r length(tar_read(cv))` cross-validation models, 
`r sum(tar_read(cv))` models (`r printnum(mean(tar_read(cv))*100, digits = 0)`%) 
continued to predict the target species as having a median posterior probability 
of tool use that was within the range of all other tool users. This 
classification rate was greater than the baseline classification rate of 
`r printnum(mean(plotSurvCure1$data %>% filter(ToolUse == "Absent") %>% pull(med) > 0.16478274) * 100, digits = 0)`% 
for species without evidence of tool use in the full model
(`r sum(plotSurvCure1$data %>% filter(ToolUse == "Absent") %>% pull(med) > 0.16478274)` 
of `r nrow(plotSurvCure1$data %>% filter(ToolUse == "Absent"))` species without 
evidence of tool use had a median posterior probability of tool use that was 
within the range of the tool-using species). Together, the area-under-the-curve 
statistic and the leave-one-species-out approach suggest that the model is able 
to adequately classify known tool users, with some error.

Figure \@ref(fig:plotSurvCure1) visualises the ranked posterior probabilities of 
tool use from the phylogenetic survival cure model for all parrot species. As 
expected, the known tool users are ranked towards the top of this list. However, 
several "tool use absent" species also rank highly on the list, despite not 
being identified as tool users in the scientific literature or in our video 
survey. In fact, according to the model, the most likely tool user is a species 
for which tool use is unobserved in our data: the blue-eyed cockatoo (_Cacatua 
opthalmica_). This species is endemic to Papua New Guinea and is relatively 
understudied, with only `r tar_read(d)$EX.Pre.Tool.Papers[165]` published papers 
and `r tar_read(d)$SearchHits[165]` video search hits, which is fewer than the 
model expects are necessary to discover tool use when it is present (Figure 
\@ref(fig:plotSurvCure2)). This species is also found in the _Cacatua_ genus, a 
clade containing several known tool users. This prediction makes sense given the 
high phylogenetic signal for tool use reported by the model (Supplementary 
Figures \@ref(fig:plotSurvCure5) and \@ref(fig:plotSurvCure6)). Beyond the 
blue-eyed cockatoo, other highly ranked species without observed evidence of 
tool use are the Meyer's parrot (_Poicephalus meyeri_), the golden parakeet 
(_Guaruba guarouba_), the long-billed corella (_Cacatua tenuirostris_), the 
Solomons cockatoo (_Cacatua ducorpsii_), the red-fronted parrot (_Poicephalus 
gulielmi_), the Cape parrot (_Poicephalus robustus_), the yellow-eared parrot 
(_Ognorhynchus icterotis_), the red-vented cockatoo (_Cacatua haematuropygia_), 
and the gang-gang cockatoo (_Callocephalon fimbriatum_). Figure 
\@ref(fig:plotPhylo1) plots these species on the parrot phylogeny, using the top 
ten highest ranked species without observed evidence of tool use as an arbitrary 
cutoff for visualisation purposes.
